\usemodule[pycon-2014]
\starttext


\section[scrum-for-teams---agata-maron]{SCRUM for Teams - Agata Maron}

Imagine teams in which projects are done on time. Problems are
communicated early, so they're solved quickly. People are engaged in
their work and loyal. Teams are self-organized and self-starting.

These are the markings of teams that have adopted agile. There are
amazing benefits to agile management practices, and at STX Next, we have
experience transitioning dozens of teams to a fast and efficient agile
development cycle.

SCRUM for Teams is practical workshop for everybody who wants to feel
and understand pure idea of working using agile methods. We gonna
experience not only being a real scrum team but also taking a different
roles like Product Owner, Scrum Master or Devs.

Whole workshop is organized like a big game-challenge. Every scrum team
will have to make the best sprint for the goal. After that exactly like
in normal scrum team we will have time for retrospective, discuss and
starting another sprint. At the end teams who win also will take prizes
for that! So we want to combine learning, working and just simple having
fun. Challenging is always the most interesting!!

\subsection[for-who]{For who?}

This workshop is for everybody really open minded and for those who want
to learn and experience new things. In favorable atmosphere we will
check how scrum really works and depends on us.

We are waiting also for people who are no experienced in scrum because
for them it can be beginning for improving projects managing. During the
workshop they will learn how to use continuous improvement and small
scrum things in everyday work.

But even If You are experienced in SCRUM this workshop is also for You.
We believe that every practical experiment of scrum working can be
valuable and for sure will give you new thoughts and ideas about agile.
So if you feel good in scrum just come and share your experience to
others and be ready for something new, which can always happen.

\subsection[how]{How?}

During the workshop we will work in scrum teams (5-8 people). Every team
will have to produce working stuff at the end of every sprint.
Challenging can be a fact that teams will compete each other. At the end
of the sprint team will draw conclusions and start another sprint.
During game we will also change roles so there will be possibility to be
a Scrum Master, Product Owner and Developer.

Additional values of that game are possibilities to share experience and
methods providing scrum teams. We will talk more about best solutions
for every team and discuss mistakes, which teams made during the game.

\subsection[results]{Results?}

At the end of the workshop every participant will know what does really
mean working in scrum team. We will create also full experience to feel
what does really sprint should work and how to start providing projects
using agile methods step by step. Like in good kaizen we give you some
ideas to start improving your project managing from the next day after
workshop.

Additional values from workshop? We know that agile is also for sharing
passion and simply having fun. So we want to integrate and spending nice
time together during the conference!

You are more than welcome to join \quotation{SCRUM for Teams}


\stoptext
