\usemodule[pycon-2015]
\starttext

\Title{Business logic with PostgreSQL and Python}
\Author{Hubert Piotrowski}
\MakeTitlePage

\subsection[introduction]{Introduction}

In my opinion PostgreSQL is the most advanced open source database. It
has so many powerful features, that it is really difficult to put them
all in one article.

In this article I'm going to show you the least known features of
PostgreSQL, which is procedural languages and stored functions. By using
stored procedures, we are going to build business logic step by step.

\subsection[business-logic]{Business logic}

By business logic, we are going to call all of those blocks of code,
which are responsible for making calculations and taking decisions about
data flow. In many cases most of developers decide to build business
logic as part of ORM models, which is pretty convenient. Whenever data
is being used from a database, ORM can take control and decide how data
is going to be returned. Some ORM frameworks, like Django, allow us to
extend base model and start building our own logic in a very complex
way.

Example piece of pseudo code in Django ORM

\starttyping
from django.db import models

class _MyBaseModel(models.Model):
  def save(self, *args, **kwargs):
    # we can have some magic here
    # code code code

    # call actual save, so "magic" will happen before saving data into DB
    super(_MyBaseModel, self).save(*args, **kwargs)

class History(_MyBaseModel):
  """ model which always calls custom save method before calling Django's save """
  shops = models.IntegerField('Number of shops', default = 0, null = True)
  customers = models.IntegerField('Number of customers', default = 0, null=True)
  customer_type = models.SmallIntegerField('Customer Type', choices = CustomerTypes.CHOICES)

  def save(self, *args, **kwargs):
      # also we can have another level of customization here
      super(History, self).save(*args, **kwargs)
\stoptyping

In this simple example, you can see how easy it is to customize save
method in Django ORM. With this custom code you can put some business
logic in the save method. For example, you can validate data prior to
saving in a database. The options here are unlimited.

Other developers prefer to represent business logic as a bunch of
classes. That is also a good solution, as long as you pay attention and
these classes are the single source of truth.

\subsection[single-source-of-truth]{Single source of truth}

To explain what I call the {\em single source of truth} I will use an
example. Imagine a web application that uses database as data storage
and it accepts some page form being posted, but in different views.
Let's assume that aforementioned form can have slight differences in
those views. What you expect from your system based on given input,
(submitted page form) is to always validate and process data in the
exact same way. For example, your system is going to check if there is
any duplication in database, validate required data and data types. It
is obvious that system always has to react in the same way and return
designed result, based on a given input.

That is what I am going to use in this article as the {\em single source
of truth}, in other words {\em Business Logic}. Each time input is
given, system will react in the same way. Each entry point of the
system, which uses the same context of data will react in the same way.

\subsection[why-not-orm]{Why not ORM?}

Using ORM as a business logic container definitely has a lot of pros,
although there is one serious problem with it. If you have a project
which uses DB that is shared with other projects and those use different
languages\ldots{}then\ldots{} you mostly are going to use API for such
interactions with third-party software projects. What if you have to
deal with legacy code, where having API is not an option? Where to store
your single source of truth, iow. your business logic? How about using
the database as one.

\subsection[python]{Python}

Compiling Python is not a difficult process. Let me show you, in a few
simples steps, how to do it.

Download python source code

\starttyping
wget https://www.python.org/ftp/python/2.7.10/Python-2.7.10.tgz
\stoptyping

Compile Python under your custom directory ({\em ---prefix} flag). In
the following example I'm going to compile Python under {\em /opt/py} to
make sure that Python which I will use later on, with PostgreSQL is not
conflicting with Python, that is installed with operating system. Custom
Python also has got one significant advantage. If operating system
(e.g.~Linux) comes with Python that is main part of system tools
(e.g.~yum) it is always good idea to isolate Python that you're about to
use with your application from system's Python.

Please make sure to add {\em --enable-shared} flag during compilation.
This option will tell Python to compile with shared libraries. Once
Python libraries are compiled with shared option then any software can
soft link them and use Python.

\starttyping
~/stuff/Python-2.7.10% ./configure —prefix=/opt/py –enable-shared
\stoptyping

When compilation is finished, some operating systems, when you try to
run your compiled Python can return the following error message:

\starttyping
py ➤ bin/python2.7

bin/python2.7: error while loading shared libraries: libpython2.7.so.1.0: cannot open shared object file: No such file or directory**
\stoptyping

You have to add {\em lib path} of your newly compiled Python to the
system lib path. For instance you can add below line to your .bashrc
file. In my case, I compiled Python under {\em /opt/py} so that is why I
added path as below:

\starttyping
export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:/opt/py/lib
\stoptyping

To simplify installation process of Python modules let's install
{\bf pip package manager}.

\starttyping
wget https://bootstrap.pypa.io/get-pip.py | /opt/py/bin/python
\stoptyping

\subsection[postgresql-with-python-support]{PostgreSQL with Python
support}

Download source code.

\starttyping
wget https://ftp.postgresql.org/pub/source/v9.4.4/postgresql-9.4.4.tar.bz2
\stoptyping

Compile with Python support from {\em /opt/py} directory. Same as during
Python compilation I'm going to compile PostgreSQL under custom
directory ({\em --prefix} flag) {\em /opt/pgsql}.

\starttyping
./configure –prefix=/opt/pgsql —with-python PYTHON=/opt/py/bin/python

make

make install
\stoptyping

Now you have to compile and install PostgreSQL plugins

\starttyping
cd contrib

make

make install
\stoptyping

\subsection[create-db]{Create DB}

Now it is time to start PostgreSQL. First you have to initialize name
space. I used {\em /opt/pg\letterunderscore{}data} directory for
PostgreSQL storage.

\starttyping
mkdir /opt/pg_data
/opt/pgsql/bin/initdb /opt/pg_data
\stoptyping

and start DB. Next command will start PostgreSQL, which is going to
listen on all interfaces and port {\em 5432} (default). Data directory
({\em -D} option) is going to be /opt/pg\letterunderscore{}data.

\starttyping
/opt/pgsql/bin/postmaster -p 5432 -D /opt/pg_data/
\stoptyping

Create new empty database {\em pie} with encoding {\bf UTF-8} ({\em -E}
option). Option {\em -h} tells createdb command that PostgreSQL server
listens on {\em localhost} and port {\em 5432} (option {\em -p}).

\starttyping
/opt/pgsql/bin/createdb -h localhost -p5432  -E utf8 pie
\stoptyping

Install Python support for database {\em pie}. Again we have to specify
localhost (-h) and port (-p) that is being used by the server. Option
{\em -d} allow you to create {\bf plpythonu} language under {\em pie}
database. Please notice, that language that I created in the following
example is called *{\em plpythonu}. Letter {\em u} stands for untrusted.
Why untrusted? That is more about history of PostgresQL and Python
support.

\starttyping
/opt/pgsql/bin/createlang -h localhost -p5432 -d pie plpythonu
\stoptyping

And that's it. Simple as that. Your new DB has full Python support and
we can start organizing business logic in there.

\subsection[word-of-the-day]{Word of the day}

Before creating your first plpython function you need to know how it
works. Your compiled Python and its modules are fully accessible from
plpython. This means that the entire Python standard library is fully
accessible, when writing your business logic. You can also install any
kind of module which you may use later from your functions.

I need to provide additional clarifications here, both terms
\quote{procedural} and \quote{function} can be a little bit confusing
for newcomers. Generally from PostgreSQL perspective you always create
plpython function. You call plpython function which is stored in DB.
What is \quotation{inside} of that function (Python objects or other
functions) is up to you to decide. Procedural programming is not the
only pattern that you can use inside of plpython. To illustrate two
different bodies of function please check example below.

\starttyping
create or replace function my_first_function()
returns int as
$$

class MyObjectA:
    def __init__(self):
        """some body of constructor"""

    def methodA(self, param_1, param_2):
        """body of the method"""

class MyObjectB:
    def __init__(self):
        """some body of constructor"""

    def methodB(self, param_1, param_2):
        """body of the method"""

b = methodB()
a = MyObjectA()
plpy.info("Some info message")
return b.methodB(a.methodA(10, 15), "some string")


$$
LANGUAGE plpythonu VOLATILE;
\stoptyping

Example of calling such a plpython function

\starttyping
sql: SELECT * FROM my_first_function();
\stoptyping

Same plpython function but with much simpler body

\starttyping
create or replace function my_first_function()
returns int as
$$

import random

return random.randint(1,20)

$$
LANGUAGE plpythonu VOLATILE;
\stoptyping

Example of calling such a plpython function

\starttyping
sql: SELECT * FROM my_first_function();
\stoptyping

You may notice that from PostgreSQL engine perspective calling
{\em my\letterunderscore{}first\letterunderscore{}function} looks
identical, even if the actual body of function written in plpython is
either flat function or complex object.

As for the business logic with Python inside of database that I want
show you, I am not trying to convince you to start using procedural
programming against object programming or the other way around. As I
already said - it is up to you to decide what will fit best when writing
plpython functions, although please read below Zen of Python. Again,
again, and\ldots{} again before you start writing actually any Python
code.

\startblockquote
Beautiful is better than ugly.
\stopblockquote

\startblockquote
Explicit is better than implicit.
\stopblockquote

\startblockquote
Simple is better than complex.
\stopblockquote

\startblockquote
Flat is better than nested.
\stopblockquote

\startblockquote
Readability counts.
\stopblockquote

\startblockquote
If the implementation is hard to explain, it's a bad idea
\stopblockquote

What I am trying to point out here is, you should be careful when
creating plpython functions, they should not be too complex. Simplicity
can bring you a lot of benefits, especially when debugging the code to
nail a bug.

\section[hello-world]{Hello world}

Let me show you, how to write the classical hello world function

\starttyping
create or replace function logic.hello_world()
returns void as
$$
"""
Code code code code
"""
plpy.info("hello world")
$$
LANGUAGE plpythonu VOLATILE;
\stoptyping

As you can see from above example, you are able to print out some log
messages. They go to stdout of PostgreSQL directly, which means in
production stdout is redirected to a log file. Depending on PostgreSQL
log level you can hide some unwanted messages from plPy function same as
some context like which function calls which. Example of a verbose log
level.

\starttyping
pie=# select * from logic.view_and_set_discounted_sales(20, 23) ;
INFO:  [view_and_set_discounted_sales] Updating item id: 36 with new discounted price: 0.529331946972
CONTEXT:  PL/Python function "view_and_set_discounted_sales"
INFO:  [view_and_set_discounted_sales] Updating item id: 40 with new discounted price: 5.03912459204
CONTEXT:  PL/Python function "view_and_set_discounted_sales"
INFO:  [view_and_set_discounted_sales] Updating item id: 45 with new discounted price: 3.37963644177
CONTEXT:  PL/Python function "view_and_set_discounted_sales"
\stoptyping

PostgreSQL with log level (notice) will not only show you your
{\em info} messages but also {\bf context} in which it's been logged.
That is something that is very helpful with debugging production cases.
Once function is being called from context statement you can see what
kind of triggers or sub-functions were called. Example message

\starttyping
INFO:  [view_and_set_discounted_sales] Updating item id: 45 with new discounted price: 3.37963644177
CONTEXT:  PL/Python function "view_and_set_discounted_sales"
\stoptyping

Let's stop here for a minute and analyze that example.

{\em INFO} - log level of the message
{\em {[}view\letterunderscore{}and\letterunderscore{}set\letterunderscore{}discounted\letterunderscore{}sales{]}
Updating item id: 45 with new discounted price: 3.37963644177} - Custom
message that function prints out {\em CONTEXT: PL/Python function
\quotation{view\letterunderscore{}and\letterunderscore{}set\letterunderscore{}discounted\letterunderscore{}sales}}
- context in which above message has been logged.

To be able to control {\bf log level} that PostgreSQL is going to catch
or ignore please use one of below options in /opt/pgsql/postgresql.conf
settings file.

{\bf client\letterunderscore{}min\letterunderscore{}messages} - Controls
which message levels are sent to the client (e.g.~PostgreSQL shell)

{\bf log\letterunderscore{}min\letterunderscore{}messages} - Controls
which message levels are written to the server log

Available values of log level in order of decreasing detail:

\startitemize[n][stopper=.]
\item
  debug5
\item
  debug4
\item
  debug3
\item
  debug2
\item
  debug1
\item
  log
\item
  notice
\item
  warning
\item
  error
\stopitemize

\subsection[database-schema]{Database schema}

First comes first. Below you can see structure of tables and relations
for database {\em pie} which I will use in future examples.

\starttyping
CREATE TABLE client
(
  id bigserial NOT NULL,
  row_change_time timestamp without time zone NOT NULL,
  c_name varchar,
  c_surname varchar,
  dob date,
  e_mail varchar,
  CONSTRAINT client_pkey PRIMARY KEY (id),
  CONSTRAINT client_email_key UNIQUE (e_mail)
);


CREATE TABLE item
(
    id bigserial NOT NULL,
    item_name text,
    item_price Decimal(10,2),
    item_serial_nr varchar(32),
    active boolean,
  CONSTRAINT item_pkey PRIMARY KEY (id),
  CONSTRAINT item_name_id_key UNIQUE (item_name, item_serial_nr)
);

CREATE TABLE bill
(
  id bigserial NOT NULL,
  bill_created timestamp without time zone NOT NULL,
  shop_code varchar(32) NOT NULL,
  field_hash varchar(32) NOT NULL,
  client_id bigint NOT NULL,
  bill_sent boolean default false,
  bill_number varchar,
  CONSTRAINT bill_pkey PRIMARY KEY (id),
  CONSTRAINT billcode_hash_key UNIQUE (shop_code, field_hash)

);

ALTER TABLE bill
  ADD FOREIGN KEY (client_id) REFERENCES client (id) ON DELETE CASCADE ON UPDATE CASCADE;



CREATE TABLE bill_item
(
    id bigserial NOT NULL,
    item_name text,
    item_qty int,
    item_price float,
    item_discount_value float default 0,
    item_id bigint,
    bill_id bigint,
  CONSTRAINT bill_item_pkey PRIMARY KEY (id),
  CONSTRAINT bill_name_id_key UNIQUE (bill_id, item_id)
);

ALTER TABLE bill_item
  ADD FOREIGN KEY (item_id) REFERENCES item (id) ON DELETE CASCADE ON UPDATE CASCADE;
ALTER TABLE bill_item
  ADD FOREIGN KEY (bill_id) REFERENCES bill(id) ON DELETE CASCADE ON UPDATE CASCADE;


CREATE SCHEMA logic;

CREATE SEQUENCE public.bill_number_seq;
\stoptyping

\subsection[main-pattern]{Main pattern}

For building business logic in database I will use two main things:

\startitemize
\item
  triggers with trigger functions on tables
\item
  plpython functions to access data from database
\stopitemize

\subsection[triggers]{Triggers}

Triggers are going to help me with making my data to be consistent. It
doesn't matter if tables are going to be controlled by ORM, or they are
going to be accessed using pure SQL. Each time data is being changed,
trigger will be used with corresponding function. Of course, based on
which conditions PostgreSQL engine will call triggered function is up to
you to define.

Trigger function can be called before or after

\startitemize
\item
  update
\item
  insert
\item
  delete
\stopitemize

How to define trigger on a table foo you can see below. Trigger function
{\em my\letterunderscore{}trigger\letterunderscore{}function} will be
called {\em after insert} of each record.

\starttyping
CREATE TRIGGER my_trigger
    AFTER INSERT ON table_foo
        FOR EACH ROW EXECUTE PROCEDURE my_trigger_function();
\stoptyping

To show you how trigger can be applied on tables let's try to look
closer on the following example. Let's create table {\em foo} and
{\em foo\letterunderscore{}backup}. Structure of these tables is as
follows:

\starttyping
CREATE TABLE foo
(
  id bigserial NOT NULL,
  row_change_time timestamp without time zone NOT NULL,
  e_mail varchar,
  CONSTRAINT foo_pkey PRIMARY KEY (id),
  CONSTRAINT foo_email_key UNIQUE (e_mail)
);
\stoptyping

To be able to protect data in table {\em foo} from being deleted let's
apply a trigger in that table.

\starttyping
CREATE TRIGGER my_trigger_before_delete
    BEFORE DELETE ON foo
        FOR EACH ROW EXECUTE PROCEDURE protect_data();
\stoptyping

I will make a copy of data when field
{\em row\letterunderscore{}change\letterunderscore{}time} is greater
then 15 minutes.

\starttyping
create or replace function protect_data()
returns trigger as
$$
from datetime import datetime

v = datetime.now() - datetime.strptime(TD['old']['row_change_time'][:19], '%Y-%m-%d %H:%M:%S')

if v.seconds/60.0 > 15:
    # backup data before losing it
    plpy.info('Backing up data...')
    sql = """INSERT INTO foo_backup (row_change_time, e_mail)
                VALUES ('{0}', '{1}')".format(TD['old']['row_change_time'],
                                            TD['old']['e_mail'])
    plpy.execute(sql)


$$
LANGUAGE plpythonu VOLATILE;
\stoptyping

\subsection[plpython-functions]{plpython functions}

As it was already mentioned, another PostgreSQL feature which is going
to be used to build business logic in a DB are {\em plpython functions}.
These functions will be used by all the applications that will access
database. Each time I will try to modify or read some data in database
{\em pie} I will call a function instead of a table directly.

Example of such approach can be plpython function which takes 2
arguments and returns set of records which are considered by PostgreSQL
as a table. Such a table works as a view, which of course can be used in
regular SQL queries. Below I pasted a function which for given bill
number will apply given discount. Just to remind you table structures
were shown in chaper "schema*.

\starttyping
create or replace function logic.view_and_set_discounted_sales(
in_bill_number bigint,
in_discount_percentage float,
out out_id bigint,
out out_bill_number varchar,
out out_item_name text,
out out_item_price Decimal(10,2),
out out_item_serial_nr varchar,
out out_bill_sent boolean
)
returns setof record as
$$

from decimal import Decimal

sql = """SELECT bill.id, bill_number, bill_item.item_name, bill_item.item_qty, bill_item.item_price, item_discount_value, bill_id, bill_sent, item_id,
item_serial_nr
FROM
bill LEFT JOIN bill_item on (bill.id =bill_id)
LEFT JOIN item on (item_id=item.id)
WHERE
bill.id=%d""" % (in_bill_number)
result = []
plpy.debug(sql);
for x in plpy.execute(sql):
    item_price = plpy.execute("SELECT item_price FROM bill_item WHERE item_id=%d AND bill_id=%d" % (x['item_id'], x['bill_id']))[0]['item_price']
    discounted_price = float(item_price)*((100.0-in_discount_percentage)/100.0)
    result.append([x['id'], x['bill_number'], x['item_name'], discounted_price, x['item_serial_nr'], x['bill_sent']])

return result
$$
LANGUAGE plpythonu VOLATILE;
\stoptyping

Executing above function on a bill table is going to return below
results. Of course I pre-filled tables with data to be easy to simulate
working function.

\starttyping
pie=# select * from logic.view_and_set_discounted_sales(20, 33);

out_id | out_bill_number |     out_item_name     | out_item_price |        out_item_serial_nr        | out_bill_sent
--------+-----------------+-----------------------+----------------+----------------------------------+---------------
 20 | 20              | My awesome product 6  | 0.687444086976 | 109a95fce2b40f146d15c743c5cd0278 | t
 20 | 20              | My awesome product 10 |    6.544317652 | 3fda3c7814fc94a0dba9fc2bb190adab | t
 20 | 20              | My awesome product 15 |  4.38913823607 | 2ad4a8800fd241d5cf6519404a385aa1 | t
(3 rows)
\stoptyping

and different value for
{\em in\letterunderscore{}discount\letterunderscore{}percentage}
parameter

\starttyping
pie=# select * from logic.view_and_set_discounted_sales(20, 23) ;

 out_id | out_bill_number |     out_item_name     | out_item_price |
--------+-----------------+-----------------------+----------------+
     20 | 20              | My awesome product 6  | 0.529331946972 |
     20 | 20              | My awesome product 10 |  5.03912459204 |
     20 | 20              | My awesome product 15 |  3.37963644177 |
|        out_item_serial_nr        | out_bill_sent
+----------------------------------+---------------
| 109a95fce2b40f146d15c743c5cd0278 | t
| 3fda3c7814fc94a0dba9fc2bb190adab | t
| 2ad4a8800fd241d5cf6519404a385aa1 | t
    (3 rows)
\stoptyping

As you can see in the above example it is possible to create function
which allows you to return dynamic table where returned data depends on
input values. Also as I said previously, PostgreSQL understands result
of such a query as a table so you can do as for instance

\starttyping
pie=# select * from logic.view_and_set_discounted_sales(20, 23)
          WHERE out_item_serial_nr='109a95fce2b40f146d15c743c5cd0278';

 out_id | out_bill_number |    out_item_name     | out_item_price |
--------+-----------------+----------------------+----------------+
     20 | 20              | My awesome product 6 | 0.407585599168 |
|        out_item_serial_nr        | out_bill_sent
+----------------------------------+---------------
| 109a95fce2b40f146d15c743c5cd0278 | t
(1 row)
\stoptyping

\subsection[simple-cache]{Simple cache}

Let's put all the mentioned triggers and plpython functions together
into one working business logic. Let's create simple caching system.

Let's start with triggers. Each time table {\em bill} is going to get
updates I will revalidate only that chunk of data that has changed and
save the result to Redis cache. Trigger definition is going to be like
this.

\starttyping
CREATE TRIGGER t_bill_i
    AFTER INSERT OR UPDATE ON bill
    FOR EACH ROW EXECUTE PROCEDURE logic.tgr_bill_i();
\stoptyping

As I mentioned before you have an access to all Python modules and
standard libraries, so let's install Redis module.

\starttyping
/opt/py/bin/pip install redis
\stoptyping

Below is the body of trigger function
{\em logic.tgr\letterunderscore{}bill\letterunderscore{}i}.

\starttyping
create or replace function logic.tgr_bill_i()
returns trigger as
$$
import redis
from cPickle import dumps
in_server = '127.0.0.1'
in_port = 6379

POOL = redis.ConnectionPool(host=in_server,
    port = in_port if in_port is not None else 6379, db = 1)
r = redis.Redis(connection_pool = POOL)

if TD['new']['bill_created']:
    bill_key = 'bill_active:%s' % TD['new']['field_hash']
    r.set(bill_key, dumps({
        'bill_created' : TD['new']['bill_created'],
        'shop_code' : TD['new']['shop_code'],
        'field_hash' : TD['new']['field_hash'],
        'id' : TD['new']['id'],
        'client_id' : TD['new']['client_id']
        }))
    plpy.info('[tgr_bill_i] Saving bill hash: {0}'.format(bill_key))
$$
LANGUAGE plpythonu VOLATILE;
\stoptyping

Once data is being saved to Redis, we can access such data by using
below plpython function.

\starttyping
create or replace function logic.get_active_bills()
returns text as
$$

import redis
from cPickle import loads
from cjson import encode

in_server = '127.0.0.1'
in_port = 6379

POOL = redis.ConnectionPool(host=in_server,
    port = in_port if in_port is not None else 6379, db = 1)
r = redis.Redis(connection_pool = POOL)

# let False means error, True - all OK
status = {'msg' : '', 'status' : False }
bills = r.keys('bill_active:*')
all_bills = []
for k in bills if bills else []:
    out = r.get(k)
    if out is None:
        continue
    data = loads(out)
    all_bills.append(data)

return encode({'status' : status, 'data' : all_bills})
$$
LANGUAGE plpythonu VOLATILE;
\stoptyping

As you may see in the above example, the function is going to return
JSON object. For instance such a data can be returned directly to a
browser without post processing (if you're writing web application).

To demonstrate you how to call above function from Python please use the
following example.

\starttyping
import psycopg2
import psycopg2.extras

def read_bills():
    conn = psycopg2.connect(my_connection_string)
    cur = conn.cursor('foo', cursor_factory=psycopg2.extras.DictCursor)

    sql = "SELECT * FROM logic.get_active_bills()"
    cur.execute(sql)
    res = cPickle.loads(conn.fetchall()[0]['get_active_bills'])

    return res # JSON structure

if __name__ == '__main__':
    print 'This is my JSON result from cache', read_bills()
\stoptyping

Most important part of the above example to notice is, that I'm
{\bf calling function} instead of making SQL query, which gets {\bf data
from table} directly. By doing so I am making sure that I actually use
my business logic. By getting data directly from a table like

\starttyping
sql = "SELECT * FROM bill"
\stoptyping

I'd bypass business logic and I could get not that kind of data which I
should get. That is actually the {\bf essence of business logic} on
database side. Calling functions and views instead of reading data
directly from tables and having bunch of triggers around tables are the
key to have business logic with PostgreSQL and Python.

\subsection[summary]{Summary}

Caching here was just an example. It also returns JSON structure instead
of set of records. The options how to transform data and what kind of
data state to return and in which format are countless. The key part, as
already mentioned, is to start using functions whenever you want to
access data. Of course, you do not have to be to extreme and
{\em always} use only functions to be able to access tables. You have to
find the right balance where it is better to allow changing data by
using functions and where data can be changed directly in the table and
triggers will do the {\em magic}.

\subsection[advantages]{Advantages}

\startitemize[n][stopper=.]
\item
  Logic on DB side makes many things simpler and much cleaner to
  control.
\item
  By having triggers you do not have to worry if data is going to be
  changed from your application, ORM, raw SQL or any third-party
  modules.
\item
  Simple to debug and replicate bugs.
\item
  Multiprocessing (controlled by database).
\item
  Direct access to data (all is running on DB side).
\item
  Clear separation from business logic and the actual code of
  application (dry code).
\stopitemize

\subsection[disadvantages]{Disadvantages}

\startitemize[n][stopper=.]
\item
  Very complex unittest.
\item
  Nested and long loops can cause significant memory foot prints.
\item
  Exclusive locks can cause altering triggers on tables (not trigger
  functions!) to hang during update unitl lock is removed.
\item
  Schema migrations always have to go in hand with functions changes.
\stopitemize

\subsection[references]{References}

\startitemize
\item
  \hyphenatedurl{http://python.org}
\item
  \hyphenatedurl{http://www.postgresql.org/docs/}
\item
  \hyphenatedurl{https://kmonsoor.wordpress.com/2015/08/10/seven-deadly-sins-in-python-code}
\stopitemize


\stoptext
